---
title: "Benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Benchmarking MLX-backed Linear and Generalized Linear Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

Rmlx::mlx_enable_compile()


library(RmlxStats)
library(Rmlx)
library(nycflights13)
library(bench)
library(fixest)
library(RcppEigen)
library(speedglm)
library(fastglm)
library(car)
library(ggplot2)
library(tidyr)
library(ggbeeswarm)
```

This vignette benchmarks RmlxStats against standard R
functions and specialized fast fitting packages (fixest, RcppEigen,
speedglm, fastglm).

## Data Preparation


```{r data-prep}
flights <- as.data.frame(nycflights13::flights)
vars <- c("arr_delay", "dep_delay", "air_time", "distance")
complete_rows <- complete.cases(flights[, vars])
bench_data <- flights[complete_rows, vars]
nrow(bench_data)
```


## Benchmark Setup

We benchmark several linear model solvers.

```{r timings, cache=FALSE}
lm_formula <- arr_delay ~ dep_delay + air_time + distance

bench_mark <- mark(
  lm = lm(lm_formula, data = bench_data),
  mlxs_lm = mlxs_lm(lm_formula, data = bench_data),
  feols = feols(lm_formula, data = bench_data),
  fastLm = RcppEigen::fastLm(lm_formula, data = bench_data),
  speedlm = speedglm::speedlm(lm_formula, data = bench_data),
  iterations = 5,
  check = FALSE
)

# Summarize median timings, memory allocation, and relative speed.
bench_summary <- data.frame(
  method = as.character(bench_mark$expression),
  median_sec = as.numeric(bench_mark$median, units = "s"),
  mem_mb = as.numeric(bench_mark$mem_alloc, units = "MB"),
  itr_per_sec = bench_mark$`itr/sec`
)
bench_summary$relative <- bench_summary$median_sec / min(bench_summary$median_sec)
bench_summary
ggplot2::autoplot(bench_mark, type = "boxplot")
```


## Agreement on Flights Benchmark

We check that different fits give close results:

```{r agreement}
lm_fit <- lm(lm_formula, data = bench_data)
mlxs_fit <- mlxs_lm(lm_formula, data = bench_data)
feols_fit <- feols(lm_formula, data = bench_data)
fastlm_fit <- RcppEigen::fastLm(lm_formula, data = bench_data)
speedlm_fit <- speedglm::speedlm(lm_formula, data = bench_data)

ref_coef <- coef(lm_fit)
ref_fitted <- fitted(lm_fit)

extract_fit <- function(coef_vec, fitted_vec) {
  c(
    max_coef = max(abs(ref_coef - coef_vec)),
    max_fitted = max(abs(ref_fitted - fitted_vec))
  )
}

fits <- list(
  mlxs = list(
    coef = as.numeric(as.matrix(mlxs_fit$coefficients)),
    fitted = as.numeric(as.matrix(mlxs_fit$fitted.values))
  ),
  feols = list(coef = coef(feols_fit), fitted = as.numeric(predict(feols_fit))),
  fastlm = list(coef = fastlm_fit$coefficients, fitted = fastlm_fit$fitted.values),
  speedlm = list(coef = speedlm_fit$coefficients, fitted = as.numeric(predict(speedlm_fit)))
)

vapply(fits, function(f) extract_fit(f$coef, f$fitted), numeric(2))
```


## High-Dimensional Benchmark


```{r highp-data}
set.seed(20251031)
n_hd <- 10000
p_hd <- 400
x_hd <- matrix(rnorm(n_hd * p_hd), nrow = n_hd, ncol = p_hd)
colnames(x_hd) <- paste0("x", seq_len(p_hd))
beta_true <- runif(p_hd, -1, 1)
y_hd <- drop(x_hd %*% beta_true + rnorm(n_hd, sd = 0.5))
hd_data <- data.frame(y = y_hd, x_hd)

options(expressions = 15000) # to avoid errors due to nested evaluation
hd_formula <- DF2formula(hd_data)
```

```{r highp-benchmark, cache=FALSE}
hd_mark <- mark(
  lm = lm(hd_formula, data = hd_data),
  mlxs_lm = mlxs_lm(hd_formula, data = hd_data),
  feols = feols(hd_formula, data = hd_data),
  fastLm = RcppEigen::fastLm(hd_formula, data = hd_data),
  speedlm = speedglm::speedlm(hd_formula, data = hd_data),
  iterations = 3,
  check = FALSE
)

hd_summary <- data.frame(
  method = as.character(hd_mark$expression),
  median_sec = as.numeric(hd_mark$median, units = "s"),
  mem_mb = as.numeric(hd_mark$mem_alloc, units = "MB"),
  itr_per_sec = hd_mark$`itr/sec`
)
hd_summary$relative <- hd_summary$median_sec / min(hd_summary$median_sec)
hd_summary
ggplot2::autoplot(hd_mark, type = "beeswarm")
```

## Agreement on the Simulated Problem


```{r highp-agreement}
lm_hd <- lm(hd_formula, data = hd_data)
mlxs_hd <- mlxs_lm(hd_formula, data = hd_data)
speedlm_hd <- speedglm::speedlm(hd_formula, data = hd_data)

ref_coef_hd <- coef(lm_hd)

vapply(
  list(
    mlxs = as.numeric(as.matrix(mlxs_hd$coefficients)),
    speedlm = speedlm_hd$coefficients
  ),
  function(coefs) max(abs(ref_coef_hd - coefs)),
  numeric(1)
)
```


## GLM Benchmark: Logistic Regression

We benchmark `mlxs_glm()` against `stats::glm()` and `speedglm::speedglm()` 
using a binomial family.

```{r glm-data-prep}
# Create binary outcome: whether arrival delay exceeds 15 minutes
bench_data$late <- as.integer(bench_data$arr_delay > 15)
glm_formula <- late ~ dep_delay + air_time + distance
```

```{r glm-timings, cache=FALSE}
glm_mark <- mark(
  glm = glm(glm_formula, family = binomial(), data = bench_data),
  mlxs_glm = mlxs_glm(glm_formula, family = mlxs_binomial(), data = bench_data),
  speedglm = speedglm::speedglm(glm_formula, family = binomial(), data = bench_data),
  iterations = 5,
  check = FALSE
)

glm_summary <- data.frame(
  method = as.character(glm_mark$expression),
  median_sec = as.numeric(glm_mark$median, units = "s"),
  mem_mb = as.numeric(glm_mark$mem_alloc, units = "MB"),
  itr_per_sec = glm_mark$`itr/sec`
)
glm_summary$relative <- glm_summary$median_sec / min(glm_summary$median_sec)
glm_summary
ggplot2::autoplot(glm_mark, type = "boxplot")
```

## Agreement on GLM Benchmark

```{r glm-agreement}
glm_fit <- glm(glm_formula, family = binomial(), data = bench_data)
mlxs_glm_fit <- mlxs_glm(glm_formula, family = mlxs_binomial(), data = bench_data)
speedglm_fit <- speedglm::speedglm(glm_formula, family = binomial(), data = bench_data)

ref_coef_glm <- coef(glm_fit)
ref_fitted_glm <- fitted(glm_fit)

glm_targets <- list(
  mlxs = list(
    coef = as.numeric(as.matrix(mlxs_glm_fit$coefficients)),
    fitted = as.numeric(as.matrix(mlxs_glm_fit$fitted.values))
  ),
  speedglm = list(
    coef = speedglm_fit$coefficients,
    fitted = as.numeric(predict(speedglm_fit))
  )
)

vapply(
  glm_targets,
  function(fit) c(
    max_coef = max(abs(ref_coef_glm - fit$coef)),
    max_fitted = max(abs(ref_fitted_glm - fit$fitted))
  ),
  numeric(2)
)
```


## High-dimensional GLM Benchmark


```{r glm-highd-data}
set.seed(20251103)
n_glm <- 5000
p_glm <- 200
x_glm <- matrix(rnorm(n_glm * p_glm), nrow = n_glm, ncol = p_glm)
colnames(x_glm) <- paste0("x", seq_len(p_glm))
beta_glm <- runif(p_glm, -0.5, 0.5)
linpred <- drop(x_glm %*% beta_glm)
prob <- 1 / (1 + exp(-linpred))
y_glm <- rbinom(n_glm, size = 1, prob = prob)

glm_hd_data <- data.frame(y = y_glm, x_glm)
glm_hd_formula <- y ~ .
```

```{r glm-highd-benchmark, cache=FALSE}
glm_hd_mark <- mark(
  glm = glm(glm_hd_formula, family = binomial(), data = glm_hd_data,
            control = list(maxit = 50)),
  mlxs_glm = mlxs_glm(glm_hd_formula, family = mlxs_binomial(), data = glm_hd_data,
                      control = list(maxit = 50, epsilon = 1e-6)),
  speedglm = speedglm::speedglm(glm_hd_formula, family = binomial(), data = glm_hd_data),
  iterations = 3,
  check = FALSE
)

glm_hd_summary <- data.frame(
  method = as.character(glm_hd_mark$expression),
  median_sec = as.numeric(glm_hd_mark$median, units = "s"),
  mem_mb = as.numeric(glm_hd_mark$mem_alloc, units = "MB"),
  itr_per_sec = glm_hd_mark$`itr/sec`
)
glm_hd_summary$relative <- glm_hd_summary$median_sec / min(glm_hd_summary$median_sec)
glm_hd_summary
ggplot2::autoplot(glm_hd_mark, type = "beeswarm")
```

## Convergence on the High-dimensional Problem


```{r glm-highd-convergence}
glm_hd_fit <- glm(glm_hd_formula, family = binomial(), data = glm_hd_data,
                  control = list(maxit = 50))
mlxs_hd_fit <- mlxs_glm(glm_hd_formula, family = mlxs_binomial(), data = glm_hd_data,
                        control = list(maxit = 50, epsilon = 1e-6))

list(
  observations = nrow(glm_hd_data),
  predictors = ncol(glm_hd_data) - 1,
  mlxs_converged = mlxs_hd_fit$converged,
  mlxs_iterations = mlxs_hd_fit$iter,
  max_coefficient_difference = max(abs(coef(glm_hd_fit) - as.numeric(as.matrix(mlxs_hd_fit$coefficients)))),
  max_fitted_difference = max(abs(fitted(glm_hd_fit) - as.numeric(as.matrix(mlxs_hd_fit$fitted.values))))
)
```



## Bootstrap Benchmarks for `summary.mlxs_lm/_glm`

For the base case bootstrap we also try `boot::boot()` with
multi-core execution, since most Apple Silicon machines expose several cores.

```{r bootstrap-bench, cache=FALSE}
set.seed(20251110)
n_boot <- 10000L
p_boot <- 50L
x_boot <- matrix(rnorm(n_boot * p_boot), nrow = n_boot)
colnames(x_boot) <- sprintf("x%02d", seq_len(p_boot))
beta_boot <- rnorm(p_boot + 1L)
y_boot <- beta_boot[1L] + x_boot %*% beta_boot[-1L] + rnorm(n_boot, sd = 0.5)
boot_data <- data.frame(y = drop(y_boot), x_boot)

fit_mlxs_boot <- mlxs_lm(y ~ ., data = boot_data)
fit_base_boot <- lm(y ~ ., data = boot_data)

boot_stat <- function(dat, idx) {
  d <- dat[idx, , drop = FALSE]
  coef(lm(y ~ ., data = d))
}

parallel_cores <- max(1L, parallel::detectCores(logical = FALSE) - 1L)

case_boot_mark <- bench::mark(
  mlxs_case = {
    sum_case <- summary(
      fit_mlxs_boot,
      bootstrap = TRUE,
      bootstrap_args = list(
        B = 50L,
        seed = 42,
        bootstrap_type = "case",
        progress = FALSE
      )
    )
    as.numeric(as.matrix(sum_case$std.error))
  },
  boot_case_serial = boot::boot(
    boot_data,
    statistic = boot_stat,
    R = 50L,
    parallel = "no"
  ),
  boot_case_parallel = boot::boot(
    boot_data,
    statistic = boot_stat,
    R = 50L,
    parallel = "multicore",
    ncpus = parallel_cores
  ),
  car_case = car::Boot(fit_base_boot, f = coef, R = 50L, method = "case"),
  iterations = 3,
  check = FALSE,
  memory = FALSE
)

resid_boot_mark <- bench::mark(
  mlxs_resid = {
    sum_resid <- summary(
      fit_mlxs_boot,
      bootstrap = TRUE,
      bootstrap_args = list(
        B = 50L,
        seed = 99,
        bootstrap_type = "residual",
        progress = FALSE
      )
    )
    as.numeric(as.matrix(sum_resid$std.error))
  },
  car_resid = car::Boot(fit_base_boot, f = coef, R = 50L, method = "residual"),
  iterations = 3,
  check = FALSE,
  memory = FALSE
)

case_summary <- data.frame(
  method = as.character(case_boot_mark$expression),
  median_sec = as.numeric(case_boot_mark$median, units = "s"),
  itr_per_sec = case_boot_mark$`itr/sec`,
  gc_per_sec = case_boot_mark$`gc/sec`
)
case_summary$relative <- case_summary$median_sec / min(case_summary$median_sec)
case_summary <- case_summary[order(case_summary$median_sec), ]

resid_summary <- data.frame(
  method = as.character(resid_boot_mark$expression),
  median_sec = as.numeric(resid_boot_mark$median, units = "s"),
  itr_per_sec = resid_boot_mark$`itr/sec`,
  gc_per_sec = resid_boot_mark$`gc/sec`
)
resid_summary$relative <- resid_summary$median_sec / min(resid_summary$median_sec)
resid_summary <- resid_summary[order(resid_summary$median_sec), ]

list(
  parallel_cores = parallel_cores,
  case = case_summary,
  residual = resid_summary
)

```{r bootstrap-plots, fig.width=7, fig.height=4.5}
plot_boot_summary <- function(df, title) {
  df$method <- factor(df$method, levels = df$method[order(df$median_sec)])
  ggplot(df, aes(x = method, y = median_sec, fill = method)) +
    geom_col(show.legend = FALSE) +
    labs(
      title = title,
      x = NULL,
      y = "Median seconds"
    ) +
    coord_flip() +
    theme_minimal(base_size = 12)
}

case_plot <- plot_boot_summary(case_summary, "Case bootstrap timing (median seconds)")
resid_plot <- plot_boot_summary(resid_summary, "Residual bootstrap timing (median seconds)")

case_plot
resid_plot
```
```


## Elastic Net on Flights Data


```{r glmnet-flights-data}
set.seed(20251103)
enet_rows <- 50000
enet_idx <- sample.int(nrow(bench_data), enet_rows)
enet_base <- bench_data[enet_idx, ]

enet_design <- model.matrix(
  ~ poly(dep_delay, 3) + poly(air_time, 3) + poly(distance, 3)
    + poly(dep_delay, 2):poly(air_time, 2)
    + poly(dep_delay, 2):poly(distance, 2)
    + poly(air_time, 2):poly(distance, 2),
  data = enet_base
)

x_enet <- enet_design[, -1, drop = FALSE]
y_enet <- enet_base$arr_delay > 15
lambda_enet <- 0.01
```

```{r glmnet-flights-benchmark, cache=FALSE}
enet_mark <- mark(
  glmnet = glmnet::glmnet(x_enet, y_enet, family = "binomial", alpha = 1,
                          lambda = lambda_enet, standardize = TRUE,
                          thresh = 1e-6, maxit = 100000),
  mlxs_glmnet = mlxs_glmnet(x_enet, y_enet, family = mlxs_binomial(), alpha = 1,
                            lambda = lambda_enet, standardize = TRUE,
                            maxit = 2000, tol = 1e-6),
  iterations = 3,
  check = FALSE
)

enet_summary <- data.frame(
  method = as.character(enet_mark$expression),
  median_sec = as.numeric(enet_mark$median, units = "s"),
  mem_mb = as.numeric(enet_mark$mem_alloc, units = "MB"),
  itr_per_sec = enet_mark$`itr/sec`
)
enet_summary$relative <- enet_summary$median_sec / min(enet_summary$median_sec)
enet_summary
ggplot2::autoplot(enet_mark, type = "beeswarm")
```

```{r glmnet-flights-agreement}
enet_ref <- glmnet::glmnet(x_enet, y_enet, family = "binomial", alpha = 1,
                           lambda = lambda_enet, standardize = TRUE,
                           thresh = 1e-6, maxit = 100000)
enet_fit <- mlxs_glmnet(x_enet, y_enet, family = mlxs_binomial(), alpha = 1,
                        lambda = lambda_enet, standardize = TRUE,
                        maxit = 2000, tol = 1e-6)

c(
  coefficients_max_difference = max(abs(as.numeric(enet_ref$beta) - enet_fit$beta[, 1])),
  intercept_difference = abs(as.numeric(enet_ref$a0) - enet_fit$a0[1])
)
```
