---
title: "Benchmarks"
output:
  rmarkdown::html_vignette:
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Benchmarks}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
params:
  develop: FALSE
---

```{r setup, include=FALSE, message=FALSE}

# To update this vignette (and the pkgdown website) run
# pkgdown::deploy_to_branch()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)

library(RmlxStats)
library(Rmlx)
library(bench)
library(fixest)
library(RcppEigen)
library(speedglm)
library(boot)
library(lmboot)
library(ggplot2)
library(huxtable)

```

We benchmark RmlxStats against base R and specialized fast fitting packages,
across varying numbers of cases (`n`) and predictors (`p`).

Benchmarking was run on an M2 Macbook Air.

## Data Generation

```{r data-generation}
set.seed(20251111)

n_sizes <- c(10000, 50000, 250000)
p_sizes <- c(50, 100, 200, 400, 800)
n_max <- max(n_sizes)
p_max <- max(p_sizes)

X <- matrix(rnorm(n_max * p_max), nrow = n_max, ncol = p_max)
colnames(X) <- paste0("x", seq_len(p_max))

beta_true <- rnorm(p_max, mean = 0, sd = 0.5)
y_continuous <- drop(X %*% beta_true + rnorm(n_max, sd = 2))

# only 1 in 10 predictors matter:
X_sparse <- X[, seq(10, p_max, 10)]
beta_true_sparse <- beta_true[seq(1, p_max, 10)]
y_sparse <- drop(X_sparse %*% beta_true_sparse + rnorm(n_max, sd = 2))

linpred <- drop(X %*% beta_true) / 5  
prob <- 1 / (1 + exp(-linpred))
y_binary <- rbinom(n_max, size = 1, prob = prob)

full_data <- data.frame(
  y_cont = y_continuous,
  y_bin = y_binary,
  y_sparse = y_sparse,
  X
)


# for fast debugging
if (params$develop) {
  p_sizes <- p_sizes/5
  n_sizes <- n_sizes/10
}

bench_grid <- expand.grid(
  n = n_sizes,
  p = p_sizes,
  stringsAsFactors = FALSE
)

bench_grid <- bench_grid[bench_grid$n > bench_grid$p, ]

all_results <- data.frame()
```

## `lm` Benchmarks

```{r lm-benchmarks}
lm_results <- list()
lm_grid <- bench_grid[bench_grid$n <= 50000 & bench_grid$p <= 200, ]

for (i in seq_len(nrow(lm_grid))) {
  n <- lm_grid$n[i]
  p <- lm_grid$p[i]

  subset_data <- full_data[1:n, c("y_cont", paste0("x", 1:p))]
  formula_str <- paste("y_cont ~", paste(paste0("x", 1:p), collapse = " + "))
  lm_formula <- as.formula(formula_str)

  bm <- mark(
    "stats::lm" = lm(lm_formula, data = subset_data),
    "RmlxStats::mlxs_lm" = {
      l <- mlxs_lm(lm_formula, data = subset_data)
      Rmlx::mlx_eval(l$coefficients)
    },
    "fixest::feols" = feols(lm_formula, data = subset_data),
    "RcppEigen::fastLm" = RcppEigen::fastLm(lm_formula, data = subset_data),
    "speedglm::speedlm" = speedglm::speedlm(lm_formula, data = subset_data),
    iterations = 3,
    check = FALSE,
    filter_gc = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "lm"
  lm_results[[i]] <- bm
}

lm_df <- do.call(rbind, lm_results)
all_results <- rbind(all_results, lm_df)
```

## `glm` Benchmarks

```{r glm-benchmarks}
glm_results <- list()
glm_grid <- bench_grid[bench_grid$n <= 50000 & bench_grid$p <= 200, ]

for (i in seq_len(nrow(glm_grid))) {
  n <- glm_grid$n[i]
  p <- glm_grid$p[i]

  subset_data <- full_data[1:n, c("y_bin", paste0("x", 1:p))]
  formula_str <- paste("y_bin ~", paste(paste0("x", 1:p), collapse = " + "))
  glm_formula <- as.formula(formula_str)

  bm <- mark(
    "stats::glm" = glm(glm_formula, family = binomial(),
                             data = subset_data,
                             control = list(maxit = 50)),
    "RmlxStats::mlxs_glm" = {
      g <- mlxs_glm(glm_formula, family = mlxs_binomial(),
                        data = subset_data,
                        control = list(maxit = 50, epsilon = 1e-6))
      Rmlx::mlx_eval(g$coefficients)
    },
    "speedglm::speedglm" = speedglm::speedglm(glm_formula, family = binomial(),
                                   data = subset_data),
    iterations = 3,
    check = FALSE,
    filter_gc = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "glm"
  glm_results[[i]] <- bm
}

glm_df <- do.call(rbind, glm_results)
all_results <- rbind(all_results, glm_df)
```

## `glmnet` Benchmarks


```{r glmnet-benchmarks}
glmnet_results <- list()
glmnet_grid <- bench_grid

for (i in seq_len(nrow(glmnet_grid))) {
  n <- glmnet_grid$n[i]
  p <- glmnet_grid$p[i]

  xvars <- paste0("x", 1:p)
  x <- full_data[1:n, xvars]
  y <- full_data[1:n, "y_sparse"]

  bm <- mark(
    "glmnet::glmnet" = glmnet::glmnet(x, y, lambda = 1/1:50),
    "RmlxStats::mlxs_glmnet" = mlxs_glmnet(x, y, lambda = 1/1:50),
    iterations = 3,
    check = FALSE,
    filter_gc = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "glmnet"
  glmnet_results[[i]] <- bm
}

glmnet_df <- do.call(rbind, glmnet_results)
all_results <- rbind(all_results, glmnet_df)
```

## Bootstrap Benchmarks

For bootstrap, we use only the smaller datasets due to computational cost.

```{r bootstrap-benchmarks, warning=FALSE}
boot_results <- list()
boot_grid <- bench_grid[bench_grid$n <= 50000 & bench_grid$p <= 200, ]

for (i in seq_len(nrow(boot_grid))) {
  n <- boot_grid$n[i]
  p <- boot_grid$p[i]

  subset_data <- full_data[1:n, c("y_cont", paste0("x", 1:p))]
  formula_str <- paste("y_cont ~", paste(paste0("x", 1:p), collapse = " + "))
  boot_formula <- as.formula(formula_str)

  fit_mlxs <- mlxs_lm(boot_formula, data = subset_data)
  fit_base <- lm(boot_formula, data = subset_data)

  # Bootstrap function for boot package
  boot_stat <- function(dat, idx) {
    coef(lm(boot_formula, data = dat[idx, , drop = FALSE]))
  }

  bm <- mark(
    boot_case = boot::boot(subset_data, statistic = boot_stat,
                          R = 50L, parallel = "no"),
    lmboot_case = lmboot::paired.boot(boot_formula, data = subset_data, 
                                      B = 50L),
    lmboot_resid = lmboot::residual.boot(boot_formula, data = subset_data, 
                                         B = 50L),
    mlxs_case = {
      s <- summary(fit_mlxs, bootstrap = TRUE,
              bootstrap_args = list(B = 50L, seed = 42,
                                   bootstrap_type = "case",
                                   progress = FALSE))
      Rmlx::mlx_eval(s$std.err)
    },
    mlxs_resid = {
      s <- summary(fit_mlxs, bootstrap = TRUE,
              bootstrap_args = list(B = 50L, seed = 42,
                                   bootstrap_type = "resid",
                                   progress = FALSE))
      Rmlx::mlx_eval(s$std.err)
    },
    iterations = 3,
    check = FALSE,
    filter_gc = FALSE,
    memory = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "Bootstrap"
  boot_results[[i]] <- bm
}

boot_df <- do.call(rbind, boot_results)
all_results <- rbind(all_results, boot_df)
```


```{r prepare-viz-data, include = FALSE}

all_results$method <- as.character(all_results$expression)
all_results$median_sec <- as.numeric(all_results$median, units = "s")
all_results$is_mlxs <- grepl("mlxs", all_results$method)

# Summary statistics
all_results_clean <- all_results[!is.na(all_results$median_sec), ]
summary_results <- aggregate(
  median_sec ~ model_type + n + p + method + is_mlxs,
  data = all_results_clean,
  FUN = median
)

calculate_relative_stats <- function (data) {
  is_base <- grepl("^stats::", data$method) | grepl("^boot::", data$method) 
  data$rel_base <- if (any(is_base)) data$median_sec/data$median_sec[is_base] else NA
  data$rel_best <- data$median_sec/min(data$median_sec)
  data
}

expanded_results <- by(summary_results, ~ n + p + model_type, FUN = calculate_relative_stats)
expanded_results <- do.call(rbind, expanded_results)
```

## Summary Tables

We compare functions both against base R implementation, and against the fastest
alternative tested.


```{r display-functions, echo = FALSE}

table_benchmark <- function(data, caption) {
  data$np <- sprintf("n = %s, p = %s", data$n, data$p)

  data[, c("np", "method", "median_sec", "rel_base", "rel_best")] |> 
    setNames(c("np", "Method", "Median seconds", "Relative to base", "Relative to best")) |> 
    as_huxtable(add_colnames = FALSE) |> 
    set_caption(caption) |> 
    set_text_color(data$is_mlxs, "Method", "orange") |> 
    set_bold(data$rel_best == 1, everywhere) |> 
    set_tb_padding(1) |> 
    set_font_size(11) |> 
    column_to_header("np", bold = TRUE, italic = TRUE, top_padding = 16, bottom_padding = 4,
                     font_size = 14) |> 
    set_number_format(everywhere, 1, "%s") |> 
    set_number_format(everywhere, "Median seconds", "%#5.2g") |> 
    set_number_format(everywhere, c("Relative to base", "Relative to best"), fmt_percent()) |> 
    set_align(everywhere, -1, ".") |> 
    set_align(everywhere, 1, "left") |> 
    add_colnames() |> 
    add_footnote("Base is base R implementation in 'stats' or 'boot' packages.")
}

plot_benchmark <- function(data, title, subtitle = "Lower is better. Scales vary across columns") {
  ggplot(data, aes(x = median_sec, y = method, fill = is_mlxs)) +
    geom_col() +
    facet_grid(p ~ n, scales = "free", labeller = "label_both") +
    scale_fill_manual(
      values = c("TRUE" = "orange", "FALSE" = "grey40"),
      labels = c("TRUE" = "RmlxStats", "FALSE" = "Other"),
      name = ""
    ) +
    labs(
      title = title,
      x = "Median Time (seconds)",
      y = "Method",
      subtitle = subtitle
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "top"
    )
}
```

## `lm` Models

```{r display-lm, echo = FALSE}
lm_data <- subset(expanded_results, model_type == "lm")
table_benchmark(lm_data, "Linear Models")
plot_benchmark(lm_data, "Linear Model Benchmark: Time by Dataset Size")
```

## `glm` Models

```{r display-glm, echo = FALSE}
glm_data <- subset(expanded_results, model_type == "glm")
table_benchmark(glm_data, "General Linear Models")
plot_benchmark(glm_data, "GLM Benchmark: Time by Dataset Size")
```

## `glmnet` Models

```{r display-glmnet, echo = FALSE}
glmnet_data <- subset(expanded_results, model_type == "glmnet")
table_benchmark(glmnet_data, "Elastic Net Models")
plot_benchmark(glmnet_data, "Elastic Net Benchmark: Time by Dataset Size")
```

## Bootstraps

```{r display-bootstrap, echo = FALSE}
boot_data <- subset(expanded_results, model_type == "Bootstrap")

boot_data$method <- factor(boot_data$method, 
                           levels = c("lmboot_resid", "mlxs_resid",
                                      "lmboot_case", "boot_case", "mlxs_case"))
table_benchmark(boot_data, "Bootstrap Models")
plot_benchmark(
  boot_data,
  "Bootstrap Benchmark: Time by Dataset Size"
)
```



