---
title: "Benchmarks"
output:
  rmarkdown::html_document
editor_options:
  chunk_output_type: console
params:
  develop: FALSE
---



<details>
<summary>Setup code</summary>
```{r setup, include=FALSE, message=FALSE}

# To update this vignette (and the pkgdown website) run
# pkgdown::deploy_to_branch()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)

library(RmlxStats)
library(Rmlx)
library(bench)
library(fixest)
library(RcppEigen)
library(speedglm)
library(irlba)
library(rsvd)
library(bigstatsr)
library(boot)
library(lmboot)
library(ggplot2)
library(huxtable)

date <- Sys.Date()
commit <- system("git rev-parse --short HEAD", intern = TRUE)
branch <- system("git branch --show-current", intern = TRUE)
rmlx_version <- packageVersion("Rmlx")

```
</details>

We benchmark RmlxStats against base R and specialized fast fitting packages,
across varying numbers of cases (`n`) and predictors (`p`). We also
check accuracy.

Benchmarking was run on an M2 Macbook Air.

 Metadata    |  Metadata 
-------------|------------
Generated on | `r date` 
Commit       | `r commit`
Branch       | `r branch`
Rmlx version | `r rmlx_version`

## Benchmarking Code

### Data Generation

<details>
<summary>Code</summary>
```{r data-generation}
set.seed(20251111)

n_sizes <- c(10000, 50000, 250000) 
p_sizes <- c(50, 100, 200, 400, 800)
n_max <- max(n_sizes)
p_max <- max(p_sizes)

X <- matrix(rnorm(n_max * p_max), nrow = n_max, ncol = p_max)
colnames(X) <- paste0("x", seq_len(p_max))

X_glmnet <- matrix(rnorm(n_max * p_max), nrow = n_max, ncol = p_max)
for (j in 2:p_max) {
  X_glmnet[, j] <- 0.6 * X_glmnet[, j - 1] + sqrt(1 - 0.6^2) * X_glmnet[, j]
}
colnames(X_glmnet) <- paste0("x", seq_len(p_max))

beta_true <- rnorm(p_max, mean = 0, sd = 0.5)
y_continuous <- drop(X %*% beta_true + rnorm(n_max, sd = 5))

# only 1 in 10 predictors matter:
beta_glmnet_true <- rep(0, p_max)
beta_glmnet_true[seq(10, p_max, 10)] <- rnorm(length(seq(10, p_max, 10)), sd = 0.35)
y_sparse <- drop(X_glmnet %*% beta_glmnet_true + rnorm(n_max, sd = 5))

linpred <- drop(X %*% beta_true) / 10
prob <- 1 / (1 + exp(-linpred))
y_binary <- rbinom(n_max, size = 1, prob = prob)

full_data <- data.frame(
  y_cont = y_continuous,
  y_bin = y_binary,
  y_sparse = y_sparse,
  X
)

# for fast debugging
if (params$develop) {
  n_sizes <- n_sizes/10
  p_sizes <- p_sizes/5
}

bench_grid <- expand.grid(
  n = n_sizes,
  p = p_sizes,
  stringsAsFactors = FALSE
)

bench_grid <- bench_grid[bench_grid$n > bench_grid$p, ]

all_results <- data.frame()
all_accuracy <- data.frame()
```

</details>

<details>
<summary>Helper functions</summary>

```{r helper-functions}

relative_rmse <- function(estimate, truth) {
  scale <- sqrt(mean(truth^2))
  sqrt(mean((estimate - truth)^2)) / max(scale, 1e-12)
}

safe_relative_rmse <- function(estimate, truth) {
  estimate <- as.numeric(estimate)
  truth <- as.numeric(truth)

  if (anyNA(estimate) || anyNA(truth) ||
      any(!is.finite(estimate)) || any(!is.finite(truth))) {
    return(NA_real_)
  }

  relative_rmse(estimate, truth)
}

make_pca_fixture <- function(n, p, rank_k, noise_sd = 2) {
  scores_raw <- qr.Q(qr(matrix(rnorm(n * rank_k), nrow = n, ncol = rank_k)))
  scores_raw <- scale(scores_raw, center = TRUE, scale = FALSE)
  scores_basis <- qr.Q(qr(scores_raw))
  rotation <- qr.Q(qr(matrix(rnorm(p * rank_k), nrow = p, ncol = rank_k)))

  sdev_true <- seq(from = 2.5, to = 1.5, length.out = rank_k)
  singular_values <- sdev_true * sqrt(n - 1)
  x_signal <- scores_basis %*% diag(singular_values, nrow = rank_k) %*% 
    t(rotation)
  x <- x_signal + matrix(rnorm(n * p, sd = noise_sd), nrow = n, ncol = p)

  colnames(x) <- paste0("pcx", seq_len(p))

  list(
    x = x,
    rotation = rotation,
    sdev = sdev_true,
    rank = rank_k
  )
}

extract_pca_rotation <- function(fit) {
  if (inherits(fit, "big_SVD")) {
    fit$v
  } else {
    fit$rotation
  }
}

extract_pca_sdev <- function(fit) {
  if (inherits(fit, "big_SVD")) {
    fit$d / sqrt(nrow(fit$u) - 1)
  } else {
    fit$sdev
  }
}

projector_error <- function(estimate, truth) {
  estimate <- as.matrix(estimate)
  truth <- as.matrix(truth)
  proj_estimate <- estimate %*% t(estimate)
  proj_truth <- truth %*% t(truth)
  sqrt(sum((proj_estimate - proj_truth)^2)) / sqrt(sum(proj_truth^2))
}

pca_accuracy_score <- function(fit, truth_rotation, truth_sdev) {
  projector_error(extract_pca_rotation(fit), truth_rotation) +
    relative_rmse(
      as.numeric(extract_pca_sdev(fit)[seq_along(truth_sdev)]),
      truth_sdev
    )
}

make_bootstrap_data <- function(n, p) {
  x <- X[1:n, 1:p, drop = FALSE]
  colnames(x) <- paste0("x", seq_len(p))
  beta <- beta_true[seq_len(p)]
  sigma <- 1 + 2 * abs(x[, 1])
  y <- drop(x %*% beta + rnorm(n, sd = sigma))

  data.frame(y_boot = y, x)
}

bootstrap_oracle_se <- function(x, beta, formula, reps) {
  sigma <- 1 + 2 * abs(x[, 1])
  estimates <- replicate(reps, {
    y <- drop(x %*% beta + rnorm(nrow(x), sd = sigma))
    coef(lm(formula, data = data.frame(y_boot = y, x)))
  })
  apply(t(estimates), 2, sd)
}

extract_bootstrap_se <- function(method, fit) {
  if (identical(method, "stats::lm")) {
    return(as.numeric(fit))
  }

  if (identical(method, "boot::boot")) {
    return(apply(fit$t, 2, sd))
  }

  if (grepl("^lmboot::", method)) {
    return(apply(fit$bootEstParam, 2, sd))
  }

  as.numeric(fit$std.error)
}

```
</code>

### `mlxs_lm`

<details>
<summary>Code</summary>
```{r lm-benchmarks}
lm_results <- list()
lm_grid <- bench_grid[bench_grid$n <= n_sizes[2] & bench_grid$p <= p_sizes[3], ]
lm_accuracy <- list()

for (i in seq_len(nrow(lm_grid))) {
  n <- lm_grid$n[i]
  p <- lm_grid$p[i]

  subset_data <- full_data[1:n, c("y_cont", paste0("x", 1:p))]
  lm_formula <- reformulate(paste0("x", 1:p), response = "y_cont")
  beta_target <- beta_true[seq_len(p)]
  fitters <- list(
    "stats::lm" = function() lm(lm_formula, data = subset_data),
    "RmlxStats::mlxs_lm" = function() {
      fit <- mlxs_lm(lm_formula, data = subset_data)
      Rmlx::mlx_eval(fit$coefficients)
      fit
    },
    "fixest::feols" = function() feols(lm_formula, data = subset_data),
    "RcppEigen::fastLm" = function() RcppEigen::fastLm(
      lm_formula,
      data = subset_data
    ),
    "speedglm::speedlm" = function() speedglm::speedlm(
      lm_formula,
      data = subset_data
    )
  )

  bm <- mark(
    "stats::lm" = fitters[["stats::lm"]](),
    "RmlxStats::mlxs_lm" = fitters[["RmlxStats::mlxs_lm"]](),
    "fixest::feols" = fitters[["fixest::feols"]](),
    "RcppEigen::fastLm" = fitters[["RcppEigen::fastLm"]](),
    "speedglm::speedlm" = fitters[["speedglm::speedlm"]](),
    iterations = 3,
    check = function (r1, r2) {
      all.equal(as.vector(coef(r1)), as.vector(coef(r2)), 
                tolerance = 1e-6)
    },
    filter_gc = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "lm"
  lm_results[[i]] <- bm

  fits <- lapply(fitters, function(fit_method) fit_method())
  scores <- lapply(names(fits), function(method) {
    beta_hat <- as.numeric(coef(fits[[method]]))[-1]
    data.frame(
      model_type = "lm",
      n = n,
      p = p,
      method = method,
      accuracy = relative_rmse(beta_hat, beta_target),
      stringsAsFactors = FALSE
    )
  })
  lm_accuracy[[i]] <- do.call(rbind, scores)
}

lm_df <- do.call(rbind, lm_results)
all_results <- rbind(all_results, lm_df)
all_accuracy <- rbind(all_accuracy, do.call(rbind, lm_accuracy))
```
</details>

### `mlxs_glm`

<details>
<summary>Code</summary>
```{r glm-benchmarks}
glm_results <- list()
glm_grid <- bench_grid[bench_grid$n <= n_sizes[2] & bench_grid$p <= p_sizes[3], ]
glm_accuracy <- list()

for (i in seq_len(nrow(glm_grid))) {
  n <- glm_grid$n[i]
  p <- glm_grid$p[i]

  subset_data <- full_data[1:n, c("y_bin", paste0("x", 1:p))]
  glm_formula <- reformulate(paste0("x", 1:p), response = "y_bin")
  beta_target <- beta_true[seq_len(p)] / 5
  fitters <- list(
    "stats::glm" = function() glm(
      glm_formula,
      family = binomial(),
      data = subset_data,
      control = list(maxit = 50)
    ),
    "RmlxStats::mlxs_glm" = function() {
      fit <- mlxs_glm(
        glm_formula,
        family = mlxs_binomial(),
        data = subset_data,
        control = list(maxit = 50, epsilon = 1e-5)
      )
      Rmlx::mlx_eval(fit$coefficients)
      fit
    },
    "speedglm::speedglm" = function() speedglm::speedglm(
      glm_formula,
      family = binomial(),
      data = subset_data
    )
  )

  bm <- mark(
    "stats::glm" = fitters[["stats::glm"]](),
    "RmlxStats::mlxs_glm" = fitters[["RmlxStats::mlxs_glm"]](),
    "speedglm::speedglm" = fitters[["speedglm::speedglm"]](),
    iterations = 3,
    check = function (r1, r2) {
      all.equal(as.vector(coef(r1)), as.vector(coef(r2)), 
                tolerance = 1e-4)
    },
    filter_gc = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "glm"
  glm_results[[i]] <- bm

  fits <- lapply(fitters, function(fit_method) fit_method())
  scores <- lapply(names(fits), function(method) {
    beta_hat <- as.numeric(coef(fits[[method]]))[-1]
    data.frame(
      model_type = "glm",
      n = n,
      p = p,
      method = method,
      accuracy = relative_rmse(beta_hat, beta_target),
      stringsAsFactors = FALSE
    )
  })
  glm_accuracy[[i]] <- do.call(rbind, scores)
}

glm_df <- do.call(rbind, glm_results)
all_results <- rbind(all_results, glm_df)
all_accuracy <- rbind(all_accuracy, do.call(rbind, glm_accuracy))
```
</details>

### `mlxs_cv_glmnet`

<details>
<summary>Code</summary>
```{r glmnet-benchmarks}
glmnet_results <- list()
glmnet_grid <- bench_grid
glmnet_accuracy <- list()
glmnet_nfolds <- 3L

for (i in seq_len(nrow(glmnet_grid))) {
  n <- glmnet_grid$n[i]
  p <- glmnet_grid$p[i]

  xvars <- paste0("x", 1:p)
  x <- X_glmnet[1:n, xvars, drop = FALSE]
  y_sparse_subset <- y_sparse[1:n]
  beta_target <- beta_glmnet_true[seq_len(p)]
  stats_time <- system.time({
    fit_stats <- glmnet::cv.glmnet(
      x,
      y_sparse_subset,
      nfolds = glmnet_nfolds
    )
    as.numeric(fit_stats$lambda.min)
  })[["elapsed"]]
  mlx_time <- system.time({
    fit_mlx <- mlxs_cv_glmnet(
      x,
      y_sparse_subset,
      nfolds = glmnet_nfolds
    )
    Rmlx::mlx_eval(fit_mlx$glmnet.fit$x_center)
    Rmlx::mlx_eval(fit_mlx$glmnet.fit$x_scale)
    as.numeric(fit_mlx$lambda.min)
  })[["elapsed"]]

  bm <- data.frame(
    expression = c("glmnet::cv.glmnet", "RmlxStats::mlxs_cv_glmnet"),
    median = bench::as_bench_time(c(stats_time, mlx_time)),
    stringsAsFactors = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "glmnet"
  for (name in setdiff(names(all_results), names(bm))) {
    bm[[name]] <- NA
  }
  bm <- bm[, names(all_results), drop = FALSE]
  glmnet_results[[i]] <- bm

  fits <- list(
    "glmnet::cv.glmnet" = fit_stats,
    "RmlxStats::mlxs_cv_glmnet" = fit_mlx
  )
  scores <- lapply(names(fits), function(method) {
    beta_hat <- as.numeric(coef(fits[[method]], s = "lambda.min"))[-1]
    data.frame(
      model_type = "glmnet",
      n = n,
      p = p,
      method = method,
      accuracy = safe_relative_rmse(beta_hat, beta_target),
      stringsAsFactors = FALSE
    )
  })
  glmnet_accuracy[[i]] <- do.call(rbind, scores)
}

glmnet_df <- do.call(rbind, glmnet_results)
all_results <- rbind(all_results, glmnet_df)
all_accuracy <- rbind(all_accuracy, do.call(rbind, glmnet_accuracy))
```
</details>

### `mlxs_prcomp`

<details>
<summary>Code</summary>
```{r pca-benchmarks}
pca_results <- list()
pca_grid <- bench_grid[bench_grid$n <= n_sizes[2] & bench_grid$p <= p_sizes[4], ]
pca_accuracy <- list()

for (i in seq_len(nrow(pca_grid))) {
  n <- pca_grid$n[i]
  p <- pca_grid$p[i]

  rank_k <- min(8L, max(4L, floor(min(n, p) / 2)))
  pca_fixture <- make_pca_fixture(n, p, rank_k)
  x <- pca_fixture$x
  x_big <- bigstatsr::FBM(nrow = n, ncol = p, init = as.matrix(x))
  rotation_true <- pca_fixture$rotation
  fitters <- list(
    "stats::prcomp" = function() prcomp(
      x,
      center = TRUE,
      scale. = FALSE,
      rank. = rank_k
    ),
    "irlba::prcomp_irlba" = function() irlba::prcomp_irlba(
      x,
      n = rank_k,
      center = TRUE,
      scale. = FALSE
    ),
    "rsvd::rpca" = function() rsvd::rpca(
      x,
      k = rank_k,
      center = TRUE,
      scale = FALSE
    ),
    "bigstatsr::big_randomSVD" = function() bigstatsr::big_randomSVD(
      x_big,
      fun.scaling = bigstatsr::big_scale(center = TRUE, scale = FALSE),
      k = rank_k
    ),
    "RmlxStats::mlxs_prcomp" = function() {
      fit <- mlxs_prcomp(
        x,
        center = TRUE,
        scale. = FALSE,
        rank. = rank_k,
        n_iter = 2,
        seed = 1
      )
      Rmlx::mlx_eval(fit$rotation)
      if (!is.null(fit$x)) {
        Rmlx::mlx_eval(fit$x)
      }
      fit
    }
  )

  bm <- mark(
    "stats::prcomp" = fitters[["stats::prcomp"]](),
    "irlba::prcomp_irlba" = fitters[["irlba::prcomp_irlba"]](),
    "rsvd::rpca" = fitters[["rsvd::rpca"]](),
    "bigstatsr::big_randomSVD" = fitters[["bigstatsr::big_randomSVD"]](),
    "RmlxStats::mlxs_prcomp" = fitters[["RmlxStats::mlxs_prcomp"]](),
    iterations = 3,
    check = FALSE,
    filter_gc = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "pca"
  pca_results[[i]] <- bm

  fits <- lapply(fitters, function(fit_method) fit_method())
  scores <- lapply(names(fits), function(method) {
    data.frame(
      model_type = "pca",
      n = n,
      p = p,
      method = method,
      accuracy = pca_accuracy_score(
        fits[[method]],
        rotation_true,
        pca_fixture$sdev
      ),
      stringsAsFactors = FALSE
    )
  })
  pca_accuracy[[i]] <- do.call(rbind, scores)
}

pca_df <- do.call(rbind, pca_results)
all_results <- rbind(all_results, pca_df)
all_accuracy <- rbind(all_accuracy, do.call(rbind, pca_accuracy))
```
</details>

### Bootstrap

For bootstrap, we use only smaller datasets due to computational cost.

<details>
<summary>Code</summary>
```{r bootstrap-benchmarks, warning=FALSE}
boot_results <- list()
boot_grid <- bench_grid[bench_grid$n <= n_sizes[2] & bench_grid$p <= p_sizes[2], ]
boot_accuracy <- list()
bootstrap_reps <- if (params$develop) 50L else 100L
oracle_reps <- if (params$develop) 50L else 200L

for (i in seq_len(nrow(boot_grid))) {
  n <- boot_grid$n[i]
  p <- boot_grid$p[i]

  subset_data <- make_bootstrap_data(n, p)
  x <- as.matrix(subset_data[paste0("x", seq_len(p))])
  beta_target <- beta_true[seq_len(p)]
  formula_str <- paste("y_boot ~", paste(paste0("x", 1:p), collapse = " + "))
  boot_formula <- as.formula(formula_str)
  fit_mlxs <- mlxs_lm(boot_formula, data = subset_data)
  oracle_se <- bootstrap_oracle_se(x, beta_target, boot_formula, oracle_reps)
  oracle_se <- oracle_se[-1]

  boot_stat <- function(dat, idx) {
    coef(lm(boot_formula, data = dat[idx, , drop = FALSE]))
  }
  fitters <- list(
    "boot::boot" = function() boot::boot(
      subset_data,
      statistic = boot_stat,
      R = bootstrap_reps,
      parallel = "no"
    ),
    "lmboot::paired.boot" = function() lmboot::paired.boot(
      boot_formula,
      data = subset_data,
      B = bootstrap_reps
    ),
    "lmboot::residual.boot" = function() lmboot::residual.boot(
      boot_formula,
      data = subset_data,
      B = bootstrap_reps
    ),
    "mlxs_case" = function() {
      summary(
        fit_mlxs,
        bootstrap = TRUE,
        bootstrap_args = list(
          B = bootstrap_reps,
          seed = 42,
          bootstrap_type = "case",
          progress = FALSE
        )
      )
    },
    "mlxs_resid" = function() {
      summary(
        fit_mlxs,
        bootstrap = TRUE,
        bootstrap_args = list(
          B = bootstrap_reps,
          seed = 42,
          bootstrap_type = "resid",
          progress = FALSE
        )
      )
    }
  )

  bm <- mark(
    "boot::boot" = fitters[["boot::boot"]](),
    "lmboot::paired.boot" = fitters[["lmboot::paired.boot"]](),
    "lmboot::residual.boot" = fitters[["lmboot::residual.boot"]](),
    "mlxs_case" = {
      fit <- fitters[["mlxs_case"]]()
      Rmlx::mlx_eval(fit$std.error)
      fit
    },
    "mlxs_resid" = {
      fit <- fitters[["mlxs_resid"]]()
      Rmlx::mlx_eval(fit$std.error)
      fit
    },
    iterations = 3,
    check = FALSE,
    filter_gc = FALSE,
    memory = FALSE
  )

  bm$n <- n
  bm$p <- p
  bm$model_type <- "Bootstrap"
  boot_results[[i]] <- bm

  fits <- lapply(fitters, function(fit_method) fit_method())
  scores <- lapply(names(fits), function(method) {
    se_hat <- extract_bootstrap_se(method, fits[[method]])[-1]
    data.frame(
      model_type = "Bootstrap",
      n = n,
      p = p,
      method = method,
      accuracy = relative_rmse(se_hat, oracle_se),
      stringsAsFactors = FALSE
    )
  })
  boot_accuracy[[i]] <- do.call(rbind, scores)
}

boot_df <- do.call(rbind, boot_results)
all_results <- rbind(all_results, boot_df)
all_accuracy <- rbind(all_accuracy, do.call(rbind, boot_accuracy))
```
</details>


## Results: Speed

We compare functions both against base R implementation, and against the fastest
alternative tested.

<details>
<summary>Display code</summary>
```{r display-functions, echo = FALSE}


all_results$method <- as.character(all_results$expression)
all_results$median_sec <- as.numeric(all_results$median, units = "s")
all_results$is_mlxs <- grepl("mlxs", all_results$method)

# Summary statistics
all_results_clean <- all_results[!is.na(all_results$median_sec), ]
summary_results <- aggregate(
  median_sec ~ model_type + n + p + method + is_mlxs,
  data = all_results_clean,
  FUN = median
)

calculate_relative_stats <- function (data) {
  is_base <- grepl("^stats::", data$method) | grepl("^boot::", data$method) 
  data$rel_base <- if (any(is_base)) data$median_sec/data$median_sec[is_base] else NA
  data$rel_best <- data$median_sec/min(data$median_sec)
  data
}

expanded_results <- by(summary_results, ~ n + p + model_type, FUN = calculate_relative_stats)
expanded_results <- do.call(rbind, expanded_results)

table_benchmark <- function(data, caption) {
  data$np <- sprintf("n = %s, p = %s", data$n, data$p)

  data[, c("np", "method", "median_sec", "rel_base", "rel_best")] |> 
    setNames(c("np", "Method", "Median seconds", "Rel. to base", "Rel. to best")) |> 
    as_huxtable(add_colnames = FALSE) |> 
    set_caption(caption) |> 
    set_text_color(data$is_mlxs, "Method", "orange") |> 
    set_bold(data$rel_best == 1, everywhere) |> 
    set_tb_padding(1) |> 
    set_font_size(11) |> 
    column_to_header("np", bold = TRUE, italic = TRUE, top_padding = 16, bottom_padding = 4,
                     font_size = 14) |> 
    set_number_format(everywhere, 1, "%s") |> 
    set_number_format(everywhere, "Median seconds", "%#5.2g") |> 
    set_number_format(everywhere, c("Rel. to base", "Rel. to best"), 
                      fmt_percent()) |> 
    set_align(everywhere, -1, ".") |> 
    set_align(everywhere, 1, "left") |> 
    add_colnames() |> 
    set_align(1, everywhere, "left") |> 
    add_footnote("Base is base R implementation in 'stats' or 'boot' packages.")
}

plot_benchmark <- function(data, title, subtitle = "Lower is better. Scales vary across columns") {
  method_levels <- c(
    unique(data$method[data$is_mlxs]),
    unique(data$method[!data$is_mlxs])
  )
  data$method <- factor(data$method, levels = rev(method_levels))
  ggplot(data, aes(x = median_sec, y = method, fill = is_mlxs)) +
    geom_col() +
    facet_grid(n ~ p, scales = "free", labeller = "label_both") +
    scale_fill_manual(
      values = c("TRUE" = "orange", "FALSE" = "grey40"),
      labels = c("TRUE" = "RmlxStats", "FALSE" = "Other"),
      name = ""
    ) +
    labs(
      title = title,
      x = "Median Time (seconds)",
      y = "Method",
      subtitle = subtitle
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "top"
    )
}

all_accuracy$is_mlxs <- grepl("mlxs", all_accuracy$method)
accuracy_results <- by(all_accuracy, ~ n + p + model_type, FUN = function(data) {
  data$rel_best <- data$accuracy / min(data$accuracy)
  data
})
accuracy_results <- do.call(rbind, accuracy_results)

table_accuracy <- function(data, caption) {
  data$np <- sprintf("n = %s, p = %s", data$n, data$p)

  data[, c("np", "method", "accuracy", "rel_best")] |>
    setNames(c("np", "Method", "Error", "Rel. to best")) |>
    as_huxtable(add_colnames = FALSE) |>
    set_caption(caption) |>
    set_text_color(data$is_mlxs, "Method", "orange") |>
    set_bold(data$rel_best == 1, everywhere) |>
    set_tb_padding(1) |>
    set_font_size(11) |>
    column_to_header(
      "np",
      bold = TRUE,
      italic = TRUE,
      top_padding = 16,
      bottom_padding = 4,
      font_size = 14
    ) |>
    set_number_format(everywhere, 1, "%s") |>
    set_number_format(everywhere, "Error", "%#5.3g") |>
    set_number_format(everywhere, "Rel. to best", fmt_percent()) |>
    set_align(everywhere, -1, ".") |>
    set_align(everywhere, 1, "left") |>
    add_colnames() |>
    set_align(1, everywhere, "left")
}

plot_accuracy <- function(data, title, subtitle = "Lower is better. Scales vary across columns") {
  method_levels <- c(
    unique(data$method[data$is_mlxs]),
    unique(data$method[!data$is_mlxs])
  )
  data$method <- factor(data$method, levels = rev(method_levels))
  ggplot(data, aes(x = accuracy, y = method, fill = is_mlxs)) +
    geom_col() +
    facet_grid(n ~ p, scales = "free", labeller = "label_both") +
    scale_fill_manual(
      values = c("TRUE" = "orange", "FALSE" = "grey40"),
      labels = c("TRUE" = "RmlxStats", "FALSE" = "Other"),
      name = ""
    ) +
    labs(
      title = title,
      x = "Error",
      y = "Method",
      subtitle = subtitle
    ) +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "top"
    )
}
```
</details>

### `mlxs_lm`

<details>
<summary>Data table</summary>

```{r display-lm-table, echo = FALSE}
lm_data <- subset(expanded_results, model_type == "lm")
table_benchmark(lm_data, "Linear Models")
```

</details>

```{r display-lm-plot, echo = FALSE}
plot_benchmark(lm_data, "Linear Model Benchmark: Time by Dataset Size")
```

### `mlxs_glm`

<details>
<summary>Data table</summary>

```{r display-glm-table, echo = FALSE}
glm_data <- subset(expanded_results, model_type == "glm")
table_benchmark(glm_data, "General Linear Models")
```

</details>

```{r display-glm-plot, echo = FALSE}
plot_benchmark(glm_data, "GLM Benchmark: Time by Dataset Size")
```

### `mlxs_cv_glmnet`

<details>
<summary>Data table</summary>

```{r display-glmnet-table, echo = FALSE}
glmnet_data <- subset(expanded_results, model_type == "glmnet")
table_benchmark(glmnet_data, "Elastic Net Models")
```

</details>

```{r display-glmnet-plot, echo = FALSE}
plot_benchmark(glmnet_data, "Elastic Net Benchmark: Time by Dataset Size")
```

### `mlxs_prcomp`

<details>
<summary>Data table</summary>

```{r display-pca-table, echo = FALSE}
pca_data <- subset(expanded_results, model_type == "pca")
table_benchmark(pca_data, "Principal Components Models")
```

</details>

```{r display-pca-plot, echo = FALSE}
plot_benchmark(pca_data, "PCA Benchmark: Time by Dataset Size")
```

### Bootstrap

<details>
<summary>Data table</summary>

```{r display-bootstrap-table, echo = FALSE}
boot_data <- subset(expanded_results, model_type == "Bootstrap")

boot_data$method <- factor(boot_data$method, 
                           levels = c("stats::lm",
                                      "lmboot::residual.boot", "mlxs_resid",
                                      "lmboot::paired.boot", "boot::boot",
                                      "mlxs_case"))
table_benchmark(boot_data, "Bootstrap Models")
```

</details>

```{r display-bootstrap-plot, echo = FALSE}
plot_benchmark(
  boot_data,
  "Bootstrap Benchmark: Time by Dataset Size"
)
```

## Results: Accuracy

### `mlxs_lm`

Errors are calculated as the relative root mean squared error of the betas
(i.e. root mean squared error of the estimated betas compared to the true betas, divided by the root mean square of the true betas).

<details>
<summary>Data table</summary>

```{r display-lm-accuracy-table, echo = FALSE}
lm_accuracy_data <- subset(accuracy_results, model_type == "lm")
table_accuracy(lm_accuracy_data, "Linear Model Errors")
```

</details>

<details>
<summary>Graph</summary>

```{r display-lm-accuracy-plot, echo = FALSE}
plot_accuracy(lm_accuracy_data, "Linear Model Errors")
```

</details>

### `mlxs_glm`

Errors are calculated as the relative root mean squared error of the estimated betas.

<details>
<summary>Data table</summary>

```{r display-glm-accuracy-table, echo = FALSE}
glm_accuracy_data <- subset(accuracy_results, model_type == "glm")
table_accuracy(glm_accuracy_data, "GLM Errors")
```

</details>

<details>
<summary>Graph</summary>

```{r display-glm-accuracy-plot, echo = FALSE}
plot_accuracy(glm_accuracy_data, "GLM Errors")
```
</details>

### `mlxs_cv_glmnet`

Errors are calculated as the relative root mean squared error of the estimated betas, for `lambda = lambda.min`.

<details>
<summary>Data table</summary>

```{r display-glmnet-accuracy-table, echo = FALSE}
glmnet_accuracy_data <- subset(accuracy_results, model_type == "glmnet")
table_accuracy(glmnet_accuracy_data, "Elastic Net Errors")
```

</details>

<details>
<summary>Graph</summary>

```{r display-glmnet-accuracy-plot, echo = FALSE}
plot_accuracy(
  glmnet_accuracy_data,
  "Elastic Net Errors"
)
```

</details>

### `mlxs_prcomp`

PCA errors are the projector error of the rotation matrix
compared to the true data-generating rotation matrix, *plus* the relative
root mean squared error of the standard deviations. Projector error is
equivalent to measuring the principal angles between the estimated and 
true PCA subspaces.

<details>
<summary>Data table</summary>

```{r display-pca-accuracy-table, echo = FALSE}
pca_accuracy_data <- subset(accuracy_results, model_type == "pca")
table_accuracy(pca_accuracy_data, "Principal Components Errors")
```

</details>

<details>
<summary>Graph</summary>

```{r display-pca-accuracy-plot, echo = FALSE}
plot_accuracy(pca_accuracy_data, "PCA Errors")
```

</details>

### Bootstraps

Errors are calculated as the root mean squared error of the bootstrap-estimated
standard errors. We calculate the true standard error by repeatedly estimating
linear models, drawing a new dependent variable from the (known) data
generating process and calculating the standard deviation of the estimates.

<details>
<summary>Data table</summary>

```{r display-bootstrap-accuracy-table, echo = FALSE}
boot_accuracy_data <- subset(accuracy_results, model_type == "Bootstrap")
table_accuracy(boot_accuracy_data, "Bootstrap SE Errors")
```

</details>

<details>
<summary>Graph</summary>

```{r display-bootstrap-accuracy-plot, echo = FALSE}
plot_accuracy(
  boot_accuracy_data,
  "Bootstrap SE Errors"
)
```

</details>
