---
title: "Fuzz test results"
author: "David Hugh-Jones"
date: "2026-05-02"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# Fuzz tests {.tabset}

Fuzz tests live in `tests/testthat/test-fuzz-*`. We test extensively,
aiming to cover hard cases and capture a number of relevant metrics: a single
pass generates about 1000 pieces of data. We can't always binary pass/fail, 
so this page is meant to show our current performance and also how it has 
evolved over time.

The GPU is 32-bit. MLX has support for float64 but not on the GPU, and Rmlx
doesn't yet support float64 at all. That limits our precision. This page
should help us figure out how much that affects performance.

* What about diagnostics?
* What about `mlxs_cv_glmnet()`? Need to write the tests first....

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggtext)
library(glue)
library(here)
library(tidyverse)

fuzz_theme <- theme_minimal() +
  theme(
        plot.background = element_rect("grey95"),
        axis.text.y.left = element_markdown(),
        panel.background = element_rect("white"),
        legend.position = "none"
    )
set_theme(fuzz_theme)

fuzz <- read_csv(here("tests/testthat/fuzz-results/github-macos-26/fuzz-results.csv"))

latest <- function(fuzz) {
  fuzz |> filter(branch == "master", datetime_utc == max(datetime_utc)) 
}

prettify <- function(x) {
  stringr::str_to_title(gsub("_", " ", x))
}

fuzz <- fuzz |> 
  mutate(
    agg = replace_values(aggregation, "value" ~ ""),
    Measure = if_else(agg == "", target, paste0(agg, "(", target, ")")),
    Case_type = prettify(case_type),
    Scenario = paste(prettify(scenario), paste("n", n), paste("p", p), sep = " / ")
  ) 


date <- Sys.Date()
commit <- system("git rev-parse --short HEAD", intern = TRUE)
branch <- system("git branch --show-current", intern = TRUE)
rmlx_version <- packageVersion("Rmlx")

```

Metadata     |  Metadata 
-------------|------------
Generated on | `r date` 
Commit       | `r commit`
Branch       | `r branch`
Rmlx version | `r rmlx_version`

## `mlxs_lm` {.tabset}

### Deterministic tests

These tests all measure absolute difference of a statistic with a reference
value in different scenarios.

* Longley, Norris and Wamperl are NIST datasets where the reference value
  was calculated independently.
* The "Epsilon", "Large N" and "Large P" scenarios compare against `stats::lm`.
  - "Epsilon" refers to near-rank deficient scenarios where one variable
    is equal to another one plus noise times epsilon
* Maximum errors are taken over all elements of the `vcov` matrix; 
  per-case `fitted` values; or the set of `coefficient`s and their
  `standard_error`s.
  

```{r, fig.height = 7}
fuzz_lm_det <- latest(fuzz) |> 
  filter(
    suite == "mlxs-lm-deterministic",
    measure == "error"
  ) |> 
  select(-suite, -family, -term, -(nreps:bootstrap_B))

palette("Dark 2")
cols <- palette()[1:6]
names(cols) <- c("max(vcov)", "max(standard_error)", "residual_sigma",
                 "r_squared", "max(fitted)", "max(coefficient)")
fuzz_lm_det |> 
  ggplot(aes(y = Measure, x = value, color = Measure)) + 
    geom_vline(xintercept = 1e-5, linetype = "dashed") + 
    geom_point() + 
    facet_wrap(vars(Case_type, Scenario), axes = "all_x") + 
    scale_x_log10(labels = scales::label_math(format = log10)) + 
    scale_color_manual(values = cols) + 
    scale_y_discrete(labels = \(x) sprintf(
      "<span style='color:%s'>%s</span>", cols[x], x)) +
    labs(x = "Error (log scale)")
```

### Monte Carlo tests


#### Bias and confidence interval coverage

We calculate confidence interval coverage over multiple Monte Carlo repetitions
as the percentage of times the CI covered the term's true value. This also lets
us generate standard errors, and hence normal-theory 95% confidence intervals
for the true confidence interval coverage!

Similarly bias is calculated as $E(estimate - truth)$ over the Monte Carlo
repetitions, and we calculate standard errors and confidence intervals the same way.

```{r}
fuzz_lm_mc <- latest(fuzz) |> 
  filter(
    suite == "mlxs-lm-monte-carlo"
  ) |> 
  select(-suite, -family, -(alpha:noise_sd))
```

* We calculate the empirical s.e. as the standard deviation of the 
  estimates over repeated Monte Carlo runs.
* Ratios much above 1 might be worrying. (How much?)
* We have bias stats. Is there a reason we don't calculate the 
  Monte Carlo s.e.?
* Do we need to record "truth" and "estimate"? Maybe it's enough
  to look up the code, on the theory we won't change these much?
* "truth" is recorded as the baseline both for "estimate" and for
  "bias", which is true but in different senses....
* We have a huge N for the very slow bootstrap, and a tiny N for
  the fast linear model....
* Also, why *is* the bootstrap so slow?

```{r}
fuzz_lm_mc |> 
  drop_na(value_se) |> 
  mutate(
    Measure = paste0(measure, " (", term, ")"),
    value_conf_low = value - 1.96*value_se,
    value_conf_hi = value + 1.96*value_se,
    ref_line = if_else(measure=="bias", 0, 0.95)
  ) |> 
  ggplot(aes(y = term, x = value, xmin = value_conf_low, 
             xmax = value_conf_hi)) +
    geom_vline(aes(xintercept = ref_line), linetype = "dotted", 
               colour = "grey30") + 
    geom_pointrange(size = 0.1) +
    facet_grid(vars(Scenario), vars(measure), scales = "free_x", axes = "all_x") +
    labs(
      title = "Bias and 95% confidence interval coverage for Monte Carlo tests",
      subtitle = "Lines show Monte Carlo confidence intervals for coverage",
      x = "Term", y = "Value"
    )
```


#### Standard errors


We compare `mlxs_lm`'s estimated standard errors, either from 
bootstrap option or from normal theory, against the empirical standard
error, calculated as the standard deviation of the estimate over all
our Monte Carlo replications.

```{r}
bootstrap_B <- na.omit(fuzz_lm_mc$bootstrap_B)
stopifnot(n_distinct(bootstrap_B) == 1)

fuzz_lm_mc |> 
  drop_na(bootstrap_B) |> 
  filter(measure == "standard_error_ratio") |> 
  mutate(
    `s.e. type` = replace_values(source, "model" ~ "normal theory")
  ) |> 
  ggplot(aes(y = term, x = value, colour = `s.e. type`)) + 
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey30") +
    geom_point() +
    facet_wrap(vars(Scenario)) +
    labs(
      title = glue("Bootstrap ({bootstrap_B[1]} bootstraps) and normal-theory s.e. ratios for non-normal error term"),
      x = "s.e. ratio"
    ) + 
    theme(
      legend.position = "bottom"
    )
  

```


## `mlxs_glm` {.tabset}

### Deterministic tests

Here we report absolute differences with the `stats:glm` reference value.
For scenario details, look at the code in `test-fuzz-glm-det.R`.

We test gaussian, Poisson and binomial family functions, but not all
families for all scenarios.

The dashed line at $10^{-5}$ is a slightly arbitrary breakpoint. In coefficient and
s.e. space we might expect errors to be below this line.

* `case_type` just says deterministic here, different from `mlxs_lm`
  which is more useful.
* The choice of p and n is worth revisiting, maybe Codex has been ducking errors...

  
```{r}
fuzz_glm_det <- latest(fuzz) |> 
  filter(suite == "mlxs-glm-deterministic", measure == "error") |> 
  select(-suite, -case_type, -(nreps:term))
```

```{r, fig.height = 20}
cols_glm <- palette()[1:6]
names(cols_glm) <- unique(fuzz_glm_det$Measure)

fuzz_glm_det |> 
  ggplot(aes(y = Measure, x = value, colour = Measure)) + 
    geom_vline(xintercept = 1e-5, linetype = "dashed", colour = "grey30") +
    geom_point() +
    facet_grid(rows = vars(Scenario), cols = vars(family), 
               axes = "all_x") + 
    scale_x_log10(breaks = 10^c(-9, -5, -1), 
                  labels = scales::label_math(format = log10)) + 
    scale_color_manual(values = cols_glm) + 
    scale_y_discrete(labels = \(x) sprintf(
      "<span style='color:%s'>%s</span>", cols_glm[x], x)) +
    labs(
      title = "Errors of mlxs_glm statistics against glm reference",
      x = "Error (log scale)"
    ) + 
    theme(
      strip.text.y = element_text(size = 7, face = "bold")
    )
```


### Monte Carlo tests

* And here case_type is just "monte carlo"
* The rmse measure really needs a comparison with `stats::glm` to be useful,
  I think
* Is it worth evaluating the standard errors separately from CI coverage?
  Should have a similar message....
* The "full" test should generate glm bootstrap Monte Carlos....

#### Bias and confidence interval coverage

```{r}
fuzz_glm_mc <- latest(fuzz) |> 
  filter(suite == "mlxs-glm-monte-carlo") |> 
  select(-suite, -case_type, -(alpha:bootstrap_B))
```

```{r}
fuzz_glm_mc |> 
  drop_na(value_se) |> 
  mutate(
    value_conf_low = value - 1.96 * value_se,
    value_conf_hi = value + 1.96 * value_se,
    ref_line = if_else(measure == "bias", 0, 0.95),
    Measure = prettify(measure),
  ) |> 
  ggplot(aes(y = term, x = value, xmin = value_conf_low, 
             xmax = value_conf_hi)) +
    geom_vline(aes(xintercept = ref_line), linetype = "dotted", 
               colour = "grey30") + 
    geom_pointrange(size = 0.1) +
    facet_grid(vars(Scenario, family), vars(Measure), scales = "free", 
               axes = "all_x") + 
    labs(
      title = "Bias and 95% confidence interval coverage for Monte Carlo tests",
      subtitle = "Lines show Monte Carlo confidence intervals for coverage",
      x = "Value", y = "Term"
    )

```

## `mlxs_glmnet` {.tabset}

### Deterministic tests

  
```{r}
fuzz_glmnet_det <- latest(fuzz) |> 
  filter(suite == "mlxs-glmnet-deterministic") |> 
  select(-suite, -case_type, -nreps, -(rank:bootstrap_B))
```

* There's a bias measure but no confint for it, also a RMSE 
* Then prediction loss of mlxs_glmnet, glmnet and an "oracle" (which is what?)
* Then various measures/ratios against either truth or glmnet
  - "risk" delta against the oracle
  - "loss" and "loss ratio" against the glmnet reference
  - here again, a simple statistic and description might help more than these
    abstract terms
* Then precision and recall of support recovery (i.e. which variables did
  we decide were non-zero, versus the truth)
* All of this is made at fixed lambdas. So as absolute statistics, are these
  much use? Isn't this "along the path" and we care more about the final value?
  - How were the lambdas selected?
  - Why are there different lambda_idx values for different outcomes?
* Alpha is usually 1, but sometimes 0.5... er... what is alpha?


We measure precision and recall of the active set (i.e.
variables with a non-zero beta), at selected lambda values on the
fitted path.

* Precision: out of all variables identified by the fit as active, what proportion
  were truly active?
* Recall: out of all truly active variables, what proportion were correctly 
  identified?


```{r}
fuzz_glmnet_det |> 
  filter(target %in% c("support_precision", "support_recall")) |> 
  mutate(
    Target = prettify(target),
    Scenario = paste0(prettify(scenario), " (", family, ") / n = ", n, " / p = ", p),
    `Lambda index` = factor(lambda_index)
  ) |> 
  ggplot(aes(y = Scenario, x = value, colour = `Lambda index`)) +
    geom_point(position = position_dodge(width = 0.5)) +
    facet_wrap(vars(Target)) +
    theme(legend.position = "bottom")
```

We also examine quality of predictions for an out-of-sample test set.The 
loss statistic is mean squared prediction error for Gaussian fits, or mean binomial
negative log-likelihood for binomial fits. We
compare against `glmnet::glmnet()` as a reference implementation, and also the "oracle"
which is the true data-generating process (hence the best possible prediction
in expectation).

```{r, fig.height = 10}

cols_glmnet <- c("Mlx" = "orange", "Reference" = "grey50", "Oracle" = "black")
fuzz_glmnet_det |> 
  filter(measure == "loss") |> 
  # oracle loss is the same at all values of lambda so let's avoid overplotting
  filter(source != "oracle" | lambda_index == 1) |>
  mutate(
    Source = prettify(source),
    Source = fct_relevel(Source, "Mlx", "Reference", "Oracle"),
    np = paste("n =", n, "/", "p =", p)
  ) |> 
  ggplot(aes(y = Source, x = value, colour = Source, shape = factor(lambda_index))) + 
    geom_point(position = position_dodge(width = 0.5)) +
    facet_grid(rows = vars(prettify(scenario), np), cols = vars(family), 
               scales = "free_x", axes = "all_x") +
    scale_y_discrete(labels = \(x) sprintf(
      "<span style='color:%s'>%s</span>", cols_glmnet[x], x)) +
    scale_colour_manual(values = cols_glmnet, guide = "none") + 
    theme(
      legend.position = "bottom",
      legend.box = "vertical",
      strip.text.y = element_text(size = 7, face = "bold")
    ) +
    labs(
      x = "Loss",
      colour = "Source",
      shape = "Lambda index"
    )
```

### Monte Carlo tests

* These are monte carlo tests with just 5 repetitions
* Not sure what they are for, meh

```{r}
fuzz_glmnet_mc <- latest(fuzz) |> 
  filter(suite == "mlxs-glmnet-monte-carlo") |> 
  select(-suite, -case_type, -(rank:term))
```


## `mlxs_prcomp` {.tabset}

### Deterministic tests

```{r}
fuzz_prcomp_det <- latest(fuzz) |> 
  filter(suite == "mlxs-prcomp-deterministic") |> 
  select(-suite, -family, -(nreps:lambda), -bootstrap_B, -term) 
```

Scenarios are designed to test hard cases.

  * Some test the exact path, some the randomized pca path
  * They vary by n, p, rank (of the pca), true rank of the data, and by noise
    which is added to the "systematic" part of the x (so it won't be exactly
    the true rank matrix)
  
Metrics include:

* `pca_sdev` : max error compared to the `stats::prcomp` reference,
  and rmse against the reference. sdev is the standard deviations of the 
  principal components/the square roots of the eigenvalues of the covariance
  matrix.
* subspace error compared to the reference; the Frobenius distance between 
  matrices.
* reconstruction loss (of the original variables) and delta to the reference
* explained variance error, max. delta to the reference. this is also a 
  statistic from the `sdev`s.
* we record one "metamorphic" case type; not sure why just for this case
  but anyway I ignore it here as it is more "diagnostic".
* Need to think about which of these are indicating problems!

```{r, fig.height = 8}
fuzz_prcomp_det |> 
  filter(measure != "diagnostic", Measure != "reconstruction",
         case_type != "metamorphic") |> 
  mutate(
    Rank = paste0("Rank ", rank, " (true ", rank_true, ", noise ", noise_sd, ")")
  ) |> 
  ggplot(aes(y = Measure, x = value, colour = method)) + 
    geom_vline(xintercept = 1e-5, linetype = "dashed", colour = "grey30") +
    geom_point(position = position_dodge2(width = 0.5)) +
    facet_wrap(vars(Scenario, Rank), axes = "all_x", scales = "free_x", ncol = 2) +
    scale_x_log10(labels = scales::label_math(format = log10)) +
    labs(
      title = "Error metrics for `mlxs_prcomp`",
      subtitle = "Reference is `stats::prcomp`",
      x = "Error (log scale)",
      colour = "PCA Method"
    ) + 
    theme(
      legend.position = "bottom"
    )

```


### Monte Carlo tests

* These have only 12 reps which seems few. Is the Monte Carlo actually buying
  us anything here? We never use it to estimate standard errors.
* As with deterministic case, seems to struggle with "sparse dense".

```{r}
fuzz_prcomp_mc <- latest(fuzz) |> 
  filter(suite == "mlxs-prcomp-monte-carlo") |> 
  select(-suite, -case_type, -family, -(alpha:lambda), -bootstrap_B, -term)
```


```{r, fig.height = 8}
fuzz_prcomp_mc |> 
  filter(measure != "diagnostic", Measure != "mean(reconstruction)") |> 
  mutate(
    Rank = paste0("Rank ", rank, " (true ", rank_true, ", noise ", noise_sd, ")")
  ) |> 
  ggplot(aes(y = Measure, x = value)) + 
    geom_vline(xintercept = 1e-5, linetype = "dashed", colour = "grey30") +
    geom_point(position = position_dodge2(width = 0.5)) +
    facet_wrap(vars(Scenario, Rank), axes = "all_x", scales = "free_x", ncol = 2) +
    scale_x_log10(labels = scales::label_math(format = log10)) +
    labs(
      title = "Error metrics for mlxs_prcomp Monte Carlo tests",
      subtitle = "Reference is stats::prcomp",
      x = "Error (log scale)"
    )

```
